---
title: "Reproducing the LCBM panel / WES Jaccard subtypes"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducing the LCBM panel / WES Jaccard subtypes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = requireNamespace("maftools", quietly = TRUE)
)
```

## Overview

`OncoclassfieR` packages the unsupervised **Jaccard 4-subtyping** used in the
LCBM / HSYY analysis. This vignette reproduces the current results from the
datasets shipped with the package.

The two cohorts are clustered **separately** with one shared parameter set.
Panel and WES samples are not pooled because their assayed gene sets differ
too much for a joint clustering to be interpreted biologically.

| Parameter | Value |
| --- | --- |
| `max_features` | 20 (top-20 most frequently mutated genes) |
| similarity | Jaccard, `|A∩B| / |A∪B|` on the per-sample mutation gene set |
| `enhancement` | 0.6 (similarity matrix raised to the 0.6 power) |
| distance | euclidean on the similarity row profiles (`pheatmap` default, **not** `1 - Jaccard`) |
| `clustering_method` | `"ward.D2"` |
| `n_clusters` | 4 |
| `drop_empty_samples` | `TRUE` |
| seed | 1234 |
| VAF filter (upstream) | `VAF >= 0.02`, missing VAF kept |

## The data

Three real datasets ship with the package:

* `LCBM_panel_maf` — 4691 variants, 358 samples, 292 genes
  (Huashan 105 + cBioPortal/MSK-IMPACT 253).
* `LCBM_WES_maf` — 1585 variants, 157 samples, 266 genes
  (WES_SHANGHAI 64 + cellreport_WES 93).
* `LCBM_sample_annotation` — clinical annotation for all 515 samples, keyed on
  `Tumor_Sample_Barcode`.

Both MAF objects are already restricted to the 298 genes shared by the
Huashan panel and MSK-505 and already filtered to `VAF >= 0.02`. The upstream
steps that produced them are **not** part of this package: the raw four-cohort
files are not distributed, so the reproduction starts from these objects.

## Reproducing the 4-subtype split

```{r cluster}
library(OncoclassfieR)

max_features      <- 20
n_clusters        <- 4
clustering_method <- "ward.D2"
enhancement       <- 0.6
seed              <- 1234

mafs <- list(panel = LCBM_panel_maf, WES = LCBM_WES_maf)
clusters <- list()

for (grp in names(mafs)) {
  set.seed(seed)
  maf_top <- preprocess_maf_top_genes(mafs[[grp]], max_features = max_features,
                                      drop_empty_samples = TRUE)
  res <- jaccard_cluster(maf_top, n_clusters = n_clusters,
                         clustering_method = clustering_method,
                         enhancement = enhancement,
                         show_colnames = FALSE, plot = FALSE)
  clusters[[grp]] <- res$sample_clusters
  cat("\n", grp, ": ", nrow(res$sample_clusters), " samples\n", sep = "")
  print(table(res$sample_clusters$cluster))
}
```

The panel cohort gives `r paste(as.integer(table(clusters$panel$cluster)), collapse = " / ")`
samples across the four subtypes and the WES cohort gives
`r paste(as.integer(table(clusters$WES$cluster)), collapse = " / ")`, matching
the published analysis (342 -> 178 / 65 / 73 / 26 and 145 -> 28 / 71 / 28 / 18).

Samples with no mutation among the top-20 genes never enter the similarity
matrix (`drop_empty_samples = TRUE`), which is why 358 panel / 157 WES samples
produce 342 / 145 rows.

### Cluster composition

```{r composition}
for (grp in names(mafs)) {
  dat <- as.data.frame(mafs[[grp]]@data)
  cohort <- as.character(dat$cohort_name[
    match(clusters[[grp]]$Tumor_Sample_Barcode,
          as.character(dat$Tumor_Sample_Barcode))])
  cat("\n", grp, " cohort:\n", sep = "")
  print(table(cluster = clusters[[grp]]$cluster, cohort = cohort))
}
```

### The most frequently mutated genes

```{r top-genes}
lapply(mafs, function(maf) {
  counts <- sort(table(maf@data$Hugo_Symbol), decreasing = TRUE)
  head(counts, 10)
})
```

### Why four subtypes?

```{r silhouette, fig.height = 4}
set.seed(seed)
res_panel <- jaccard_cluster(
  preprocess_maf_top_genes(LCBM_panel_maf, max_features = max_features,
                           drop_empty_samples = TRUE),
  n_clusters = n_clusters, clustering_method = clustering_method,
  enhancement = enhancement, plot = FALSE)

sil <- silhouette_analysis(res_panel, k_range = 3:10,
                           clustering_method = clustering_method)
sil$silhouette
```

### The similarity heatmap

```{r heatmap, fig.width = 8, fig.height = 7}
hm <- plot_jaccard_heatmap(res_panel, group_order = as.character(1:4),
                           clustering_method = clustering_method,
                           show_colnames = FALSE, plot = FALSE)
print(hm$heatmap)
```

## Per-subtype characterisation

```{r features}
top_genes <- cluster_top_genes(LCBM_panel_maf, res_panel, top = 5)
print(top_genes)

plot_cluster_piechart(top_genes, top = 3)
```

An oncoplot of one subtype and the TMB comparison need a `maftools` `MAF`
object, which the shipped `LCBM_*.maf` objects already are:

```{r oncoplot, eval = FALSE}
plot_cluster_oncoplot(LCBM_panel_maf, res_panel, cluster_id = 1, top = 20)
```

```{r tmb, eval = FALSE}
tmb <- plot_tmb_by_cluster(LCBM_panel_maf, res_panel)
tmb$plot
tmb$wilcox
```

## Downstream: subtypes, histology and survival

The TP53 / EGFR four-group split is derived from the MAF itself, so it needs no
external data:

```{r tp53-egfr}
dat <- as.data.frame(LCBM_panel_maf@data)
tp53_mut <- unique(as.character(dat$Tumor_Sample_Barcode[dat$Hugo_Symbol == "TP53"]))
egfr_mut <- unique(as.character(dat$Tumor_Sample_Barcode[dat$Hugo_Symbol == "EGFR"]))

subtype <- res_panel$sample_clusters
subtype$TP53 <- ifelse(subtype$Tumor_Sample_Barcode %in% tp53_mut, "Mut", "WT")
subtype$EGFR <- ifelse(subtype$Tumor_Sample_Barcode %in% egfr_mut, "Mut", "WT")
subtype$group <- paste(subtype$TP53, subtype$EGFR, sep = " / ")

print(table(subtype$group))
print(table(cluster = subtype$cluster, group = subtype$group))
```

Histology comes from the annotation table. It follows the cBioPortal vocabulary
and is mapped onto the four analysis subtypes (`NET` is merged into `LCNEC`):

```{r histology}
hist_map <- c(LUAD = "AdC", LUSC = "SqCC", SCLC = "SCLC",
              LCNEC = "LCNEC", NET = "LCNEC")

anno <- LCBM_sample_annotation[
  match(subtype$Tumor_Sample_Barcode, LCBM_sample_annotation$Tumor_Sample_Barcode),
  c("Tumor_Sample_Barcode", "Histology")]
anno$subtype <- unname(hist_map[as.character(anno$Histology)])
anno$cluster <- subtype$cluster

print(table(cluster = anno$cluster, histology = anno$subtype, useNA = "ifany"))
```

Survival analysis uses `fit_km()` / `fit_cox()` on the same annotation table:

```{r survival, fig.height = 5}
cl <- res_panel$sample_clusters

surv <- LCBM_sample_annotation[
  match(cl$Tumor_Sample_Barcode, LCBM_sample_annotation$Tumor_Sample_Barcode),
  c("Tumor_Sample_Barcode", "OS_time", "OS_status")]
surv$cluster <- factor(cl$cluster)
surv <- surv[!is.na(surv$OS_time) & !is.na(surv$OS_status), ]
surv$OS_time <- as.numeric(surv$OS_time)
surv$OS_status <- as.integer(surv$OS_status)

fit <- fit_km(surv, time = "OS_time", event = "OS_status", group = "cluster")
print(fit)
```

```{r cox, eval = FALSE}
summary(fit_cox(surv, time = "OS_time", event = "OS_status", group = "cluster"))
```

```{r km, fig.height = 6, eval = FALSE}
plot_km_curves(fit, surv)
```

## Running the full reproduction

`inst/examples/LCBM_reproduction.R` runs everything in one go -- both cohorts,
the silhouette curve, the heatmaps, the per-subtype oncoplots, the TMB
comparison and the subtypes on the similarity heatmap -- and compares every
table it produces against the expected results shipped in
`inst/extdata/LCBM_expected/`:

```r
library(OncoclassfieR)
source(system.file("examples", "LCBM_reproduction.R", package = "OncoclassfieR"))
```

## Reference

`261005_panel_WES_jaccard_reproduction.md` in the analysis repository documents
the upstream chain: which four cohort objects were merged, how the 298-gene
intersection and the VAF filter were applied, and which saved result each
number corresponds to.
